---
title: "Processing VMS data in R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Preamble

This document describes how to process VMS data in R. The steps are as follows:

1. Load needed libraries and helper functions
2. Import the raw VMS-data
3. Filter the raw data
4. Grid the filtered data
5. Create a plot
6. Export the data so it can be imported into QGIS

# 1. Load the need libraries and helper functions

Needed libraries:
```{r}
library(tidyverse)
```


The helper-function we need are stored as an R-script called **functions.R** in the R-directory in the project folder. To make them "active" one needs to run the following command in R:
```{r}
source("R/functions.R")
```

# 2. Import the raw data

Note: Here we expect that the data is in a csv-format:

```{r}
file.name <- "data-raw/sierra-leone_mail-from-stephen_2015-05-14.csv"
vms.raw <- read_csv(file.name)
```

# 3. Filter the raw data

Here we want to filter the data with respect to longitudinal and latitudinal range as well as speed:

```{r}
vms.tidy <-
  vms_filter_data(vms.raw,
                  lon.min = -14.5, lon.max = -8,
                  lat.min = 6.5, lat.max = 9.5,
                  speed.min = 2.5, speed.max = 4.5)
```

# 4. Grid the filtered data

```{r}
vms.grid <-
  vms_grid_data(vms.tidy, dx = 0.005)
```

# 5. Create a plot in R

```{r}
vms_plot_data(vms.grid)
```

# 6. Export the data so it can be imported into QGIS

```{r}
write_csv(vms.grid, path = "data-raw/vms-gridded.csv")
```

